### 2024-07-01 FinalThesis/Chapter_03/Manuscript.tex
Publish PHD

\documentclass[abstract=on,10pt,a4paper,bibliography=totocnumbered]{article}
\usepackage[paper=a4paper,left=35mm,right=35mm,top=25mm,bottom=30mm]{geometry}
\usepackage[doublespacing]{setspace}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{colortbl}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{gensymb}
\usepackage{textcomp}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{enumerate}
\usepackage{enumitem}
\usepackage{subcaption}
\usepackage{booktabs}
\usepackage[hidelinks]{hyperref}
\usepackage[nameinlink]{cleveref}
\usepackage{multirow}
\usepackage{arydshln}
\usepackage[flushleft]{threeparttable}
\usepackage[nomarkers, nolists]{endfloat}
\usepackage{scalerel}
\usepackage{makecell}
\usepackage{ifthen}
\usepackage{tikz}
\usepackage{lineno}
\usetikzlibrary{svg.path}

%------------------------------------------------------------------------------
%  Bibtex vs Biblatex
%------------------------------------------------------------------------------
% While bibtex is very nice to use, it's insanely slow. I therefore work with
% bibtex until the final stage. Set the value below to "true" or "false" to pick
% which one should be used.
\newboolean{usebiblatex}
\setboolean{usebiblatex}{false}

% Depending on the question, the commands below will prepare the document
% accordingly
\ifthenelse{\boolean{usebiblatex}}{
  %-----------------------------------------------------------------------------
  %  If using biblatex
  %-----------------------------------------------------------------------------
  % Load required packages and refer to correct bibliography
  \usepackage{csquotes}
  \usepackage[
      backend=biber
    , style=apa          % To display Author-year in the text
    , natbib=true        % Necessary to use citep, etc.
    , hyperref=true      % To have clickable links in the references
    , useeditor=false    % Don't add editor to publications
    , sorting=ynt        % Sort by year, name, and title
    , uniquename=false   % Avoid that initials are put
  ]{biblatex}
  \addbibresource{Literature.bib}

  % To include author in link
  \DeclareFieldFormat{citehyperref}{%
  \DeclareFieldAlias{bibhyperref}{noformat}% Avoid nested links
  \bibhyperref{#1}}
  \DeclareFieldFormat{textcitehyperref}{%
  \DeclareFieldAlias{bibhyperref}{noformat}% Avoid nested links
  \bibhyperref{%
    #1%
    \ifbool{cbx:parens}
      {\bibcloseparen\global\boolfalse{cbx:parens}}
      {}}}
  \savebibmacro{cite}
  \savebibmacro{textcite}
  \renewbibmacro*{cite}{%
    \printtext[citehyperref]{%
      \restorebibmacro{cite}%
      \usebibmacro{cite}}}
  \renewbibmacro*{textcite}{%
    \ifboolexpr{
      ( not test {\iffieldundef{prenote}} and
        test {\ifnumequal{\value{citecount}}{1}} )
      or
      ( not test {\iffieldundef{postnote}} and
        test {\ifnumequal{\value{citecount}}{\value{citetotal}}} )
    }
      {\DeclareFieldAlias{textcitehyperref}{noformat}}
      {}%
    \printtext[textcitehyperref]{%
      \restorebibmacro{textcite}%
      \usebibmacro{textcite}}}

  % Remove editor form everything but books
  \AtEveryBibitem{%
   \ifentrytype{book}{}{%
    \clearname{editor}
   }
  }
}{
  %-----------------------------------------------------------------------------
  %  If using bibtex
  %-----------------------------------------------------------------------------
  % Load required packages and select citation style
  \usepackage[round]{natbib}
  \bibliographystyle{apalike}
}

%------------------------------------------------------------------------------
%	Some Styling
%------------------------------------------------------------------------------
% Link colors
\definecolor{linkcolor}{HTML}{000000}
\hypersetup{
  colorlinks = true,
  linkcolor  = linkcolor,
  urlcolor   = linkcolor,
  citecolor  = linkcolor,
}

% Changing the style of captions in figures etc.
\captionsetup{labelfont=bf, format=plain, font=small}

% Change how equations are referenced
\renewcommand{\theequation}{Equation \arabic{equation}}%

% Avoid skip when including text from external .tex file
\newcommand{\inputy}[1]{\input{#1}\unskip}

% For supplementary material
\newcommand{\beginappendix}{%
  \setcounter{table}{0}
  \renewcommand{\thetable}{S\arabic{table}}%
  \setcounter{figure}{0}
  \renewcommand{\thefigure}{S\arabic{figure}}%
  \setcounter{equation}{0}
  \renewcommand{\theequation}{Equation S\arabic{equation}}%
  \setcounter{section}{0}
  \renewcommand{\thesection}{A.\arabic{section}}%
}

%------------------------------------------------------------------------------
%  ORCID
%------------------------------------------------------------------------------
% Creating some TikZ styles
\tikzset{
  nonterminal/.style = {rectangle
    , minimum size = 6mm
    , very thick
    , draw = black!
  }
}

% Create the ORCID logo
\definecolor{orcidlogocol}{HTML}{A6CE39}
\tikzset{
  orcidlogo/.pic={
    \fill[orcidlogocol]
      svg{M256,128c0,70.7-57.3,128-128,128C57.3,256,0,198.7,0,128C0,57.3,57.3
        ,0,128,0C198.7,0,256,57.3,256,128z};
    \fill[white]
      svg{M86.3,186.2H70.9V79.1h15.4v48.4V186.2z}
      svg{M108.9,79.1h41.6c39.6,0,57,28.3,57,53.6c0,27.5-21.5,53.6-56.8
        ,53.6h-41.8V79.1z M124.3,172.4h24.5c34.9,0,42.9-26.5
        ,42.9-39.7c0-21.5-13.7-39.7-43.7-39.7h-23.7V172.4z}
      svg{M88.7,56.8c0,5.5-4.5,10.1-10.1,10.1c-5.6,0-10.1-4.6-10.1-10.1c0-5.6
      ,4.5-10.1,10.1-10.1C84.2,46.7,88.7,51.3,88.7,56.8z};
  }
}

% Command to create the ORCID
\newcommand\orcid[1]{\href{https://orcid.org/#1}{\mbox{\scalerel*{
\begin{tikzpicture}[yscale=-1,transform shape]
  \pic{orcidlogo};
\end{tikzpicture}
}{|}}}}

%------------------------------------------------------------------------------
%	Titlepage: Header
%------------------------------------------------------------------------------
\title{The Effects of Increasing Seasonal Dynamism when Predicting Connectivity:
Advantages or Unnecessary Complications?}

% List of Authors
\author{
  David D. Hofmann\textsuperscript{1,2,\S} \orcid{0000-0003-3477-4365} \and
  Dominik M. Behr\textsuperscript{1,2} \orcid{0000-0001-7378-8538} \and
  John W. McNutt\textsuperscript{2} \and
  Arpat Ozgul\textsuperscript{1, 2} \orcid{0000-0001-7477-2642} \and
  Gabriele Cozzi\textsuperscript{1,2} \orcid{0000-0002-1744-1940}
}

% Reduce spacing between authors
\makeatletter
\def\and{%
  \end{tabular}%
  \hskip -0.5em \@plus.17fil\relax
  \begin{tabular}[t]{c}}
\makeatother

% Current Date
% \date{\today}

% And here the masterpiece begins
\begin{document}

% Change page numbering
\pagenumbering{gobble}

% Create Titlepage
\maketitle

%------------------------------------------------------------------------------
%	Titlepage: Additional Info
%------------------------------------------------------------------------------
\begin{flushleft}

\vspace{0.5cm}

\textsuperscript{1} Department of Evolutionary Biology and Environmental
Studies, University of Zurich, Winterthurerstrasse 190, 8057 Zurich,
Switzerland.

\textsuperscript{2} Botswana Predator Conservation Program, Wild Entrust,
Private Bag 13, Maun, Botswana.

\textsuperscript{\S} Corresponding author: \href{mailto://david.hofmann2@uzh.ch}{david.hofmann2@uzh.ch}

\vspace{4cm}

\textbf{Running Title:} Connectivity in Seasonal Landscapes

\vspace{0.5cm}

\textbf{Keywords:} African wild dog, Connectivity, Dispersal, Dynamism,
Individual-based, Lycaon pictus, Okavango Delta, Seasonality

\end{flushleft}

%------------------------------------------------------------------------------
%	Abstract
%------------------------------------------------------------------------------
\newpage
\begin{abstract}

Seasonally changing conditions can drastically alter landscape connectivity.
Nevertheless, most connectivity studies ignore seasonal dynamism and instead
employ a static set of spatial covariates and assume their focal species to
exhibit a static set of preferences. Ignoring seasonality may, however, mask
important ecological features and processes, thus resulting in poor agreement
between predicted and observed movements and a misrepresentation of
connectivity.

@@

Capitalizing on natural seasonal fluctuations of the Okavango Delta in northern
Botswana and on GPS data collected on dispersing African wild dogs
(\textit{Lycaon pictus}) across different seasons, we investigate the degree to
which a better representation of seasonal dynamism improves our ability to
predict connectivity. For this, we fit integrated step-selection functions and
predict connectivity using an individual-based dispersal simulation while
explicitly considering seasonal dynamism in both environmental covariates and
the species' preferences.  Using a rigorous cross-validation procedure, we
compare the predictive model performance under each of the six proposed
configurations. While we expected that an increasing degree of seasonal dynamism
would lead to improved predictions, we were particularly interested in
identifying at which stage the inclusion of seasonality provides the biggest
benefits.

We show that, for our study system, improvements in predictive performance by
incorporating seasonal dynamism were moderate. In fact, incorporating
seasonality only improved predictions when an overly simplistic movement model
was assumed. Upon fitting a more complex model, the benefits of accounting for
seasonality vanished, resulting in imperceptible performance differences.
Despite this, patterns of connectivity as obtained from dispersal simulations
revealed marked differences between the most static and most dynamic
configurations. Most notably, connectivity was more homogeneously distributed
throughout the study area when seasonality was taken into account, suggesting
the existence of seasonal stepping stones that facilitate dispersal into
otherwise inaccessible areas.

@@

\end{abstract}

%------------------------------------------------------------------------------
%	Main Text
%------------------------------------------------------------------------------
\newpage

\onehalfspacing
\tableofcontents
\doublespacing

% Change page numbering
\newpage
\pagenumbering{arabic}

% Create linenumbers
\linenumbers
\section{Introduction}
\subsection{Connectivity}

Landscape connectivity is defined as the degree to which the landscape
facilitates or impedes movement among habitat patches \citep{Taylor.1993} and is
a critical prerequisite for maintaining biodiversity \citep{Fahrig.2003}.
Improved connectivity facilitates dispersal \citep{Doerr.2011, Baguette.2013},
which in turn promotes genetic diversity \citep{Perrin.2000, Frankham.2002} and
the colonization of vacant habitats \citep{Hanski.1999, MacArthur.2001}. Due to
its beneficial impacts on metapopulation dynamics, restoring connectivity is
among the most frequently recommended strategies to conserve biodiversity and to
promote resilience against climate change \citep{Heller.2009, Rudnick.2012}.
Quantifying connectivity and identifying critical dispersal corridors have
therefore become critical tasks in conservation science \citep{Heller.2009,
Rudnick.2012, Keeley.2019, Hofmann.2021}. One aspect that has received limited
attention in the study of connectivity is the role of seasonality. Yet, seasonal
changes both in the environment and in an animal's propensity to use or avoid
particular habitat types can profoundly impact landscape connectivity
\citep{Zeller.2020a}.

\subsection{Seasonality}

Seasonality can impact functional connectivity through spatio-temporal variation
in the landscape itself, or through temporal variation in species' preferences
towards prevailing conditions \citep{Mui.2017, Simpkins.2017a, Zeller.2020a}. In
ecosystems that experience alternations between wet and dry seasons, for
example, the onset of the rainy season initiates distinct ``green-up'' waves,
which affect the availability of food resources for herbivores and,
subsequently, shape their movements \citep{Merkle.2016}. In its most remarkable
form, the variation in environmental conditions drives herbivore migrations
across massive spatial scales, resulting in short-lived movement corridors
between two otherwise disconnected habitats \citep{Serneels.2001, Naidoo.2016}.
Alternatively, seasonality can affect functional connectivity via temporal
changes in a species' movement and habitat preferences. Amphibians, for
instance, require both aquatic and terrestrial habitats, but their preference
for one over the other heavily depends on the season \citep{Baldwin.2006}.
Although such seasonal intricacies are likely to play a fundamental role in many
ecosystems, they only rarely enter connectivity studies in an explicit manner.
In fact, most connectivity studies represent their study system by a static set
of environmental layers and assume that their focal species exhibits a fixed set
of preferences (e.g., \citealp{Elliot.2014, Abrahms.2017, Brennan.2020}).
However, this may result in biased connectivity estimates and a misallocation of
scarce conservation funds \citep{Osipova.2019, Zeller.2020}. Therefore, a more
dynamic approach to connectivity that acknowledges and renders seasonal
variation has been recommended \citep{Zeller.2020a}.

\subsection{Modeling Functional Connectivity}

Functional connectivity can be estimated using a variety of modeling techniques
\citep{Diniz.2019}, which all comprise four main steps. First, presence data of
the focal species, preferably collected during dispersal (\citealp{Elliot.2014,
Vasudev.2015, Benz.2016}, but see \citealp{Fattebert.2015}), and a set of
spatial covariate layers that are believed or known to be critical determinants
of connectivity are compiled. Second, these data are combined and fed into a
selection model that enables estimating a species selection or avoidance of
environmental features. Popular frameworks for estimating preferences are
case-control designs, where characteristics extracted at observed locations are
contrasted with characteristics extracted at available locations
\citep{Beyer.2010, Fieberg.2010}. This can be achieved using point-selection
functions \citep{Boyce.2002, Manly.2007}, path-selection functions
\citep{Cushman.2010}, and step-selection functions \citep{Fortin.2005,
Thurfjell.2014}. A particularly powerful approach is that of \textit{integrated}
step-selection functions (iSSFs), as it provides an effective means to model
both an animal's habitat-selection and movement capacity \citep{Avgar.2016,
Fieberg.2021}. Third, inferred preferences are used to predict a permeability
surface, which indicates the expected ease or difficulty at which the focal
species can traverse a certain area given the area's environmental
characteristics \citep{Zeller.2012}. Finally, in a fourth step, the permeability
surface serves as an input to a connectivity model that reveals crucial movement
corridors. At present, the most popular connectivity models are least-cost path
analysis \citep{Adriaensen.2003} and circuit theory \citep{McRae.2008}, although
individual-based movement models (IBMMs) have gained some momentum recently
\citep{Kanagaraj.2013, Allen.2016, Hauenstein.2019, Zeller.2020,
UnnithanKumar.2022a, UnnithanKumar.2022, Hofmann.2023}. In particular, some
IBMMs allow estimating connectivity directly via simulated dispersal
trajectories, thus bypassing the generation of a permeability surface
\citep{UnnithanKumar.2022a, Hofmann.2023}.

\subsection{Incorporating Seasonality in Functional Connectivity Models}

Seasonality can enter the connectivity modeling workflow described above in
three distinct stages (\Cref{GraphicalAbstract}). In the first stage, one can
either extract environmental features, such as water and vegetation, from a
static layer (i.e., a single snapshot) or from a time series of layers (i.e., a
sequence of snapshots) that capture seasonal variation across the landscape. The
former approach has historically been the norm (e.g., \citealp{Elliot.2014,
Brennan.2020}), yet advances in remote sensing technologies and a facilitated
access to petabytes of landscape data have opened up new avenues for obtaining
spatial layers at unprecedented spatio-temporal resolutions \citep{Toth.2016,
Rumiano.2020}, such that the representation of study systems via dynamic
covariate layers has become more frequent (e.g., \citealp{Osipova.2019,
Kaszta.2021}). In a second stage, one can assume their focal species to exhibit
a fixed set of preferences across seasons by pooling all presence data, or can
try to account for seasonal changes in preferences by splitting the data
accordingly (e.g., \citealp{Fortin.2005, Manly.2007, Cushman.2010,
Zeller.2020}). \cite{Chetkiewicz.2009}, for instance, partitioned their data by
season to derive seasonal habitat preferences for pumas (\textit{Puma concolor})
and grizzlies (\textit{Ursus arctos}). In a final stage, one can estimate
connectivity for either ``average'' environmental conditions, or utilize
seasonally updated layers to estimate connectivity for distinct seasons. For
connectivity models that rely on permeability surfaces, this implies repeatedly
applying the connectivity model using seasonally updated resistance surfaces
(e.g. \citealp{Osipova.2019, Zeller.2020, Kaszta.2021, Ciudad.2021}). This is
demonstrated \textit{ad extremum} by \citet{Kaszta.2021}, who prepared monthly
updated permeability surfaces for African elephants (\textit{Loxodonta
africana}). The above-mentioned IBMMs allow for an elegant solution, for
seasonality can be accounted for as simulated individuals move. Such an approach
has, however, not yet been followed. Irrespective of the method used, including
seasonality is analytically and computationally  demanding
\citep{Bishop-Taylor.2018}, raising the question to what degree it should be
considered. Our goal is (1) to create a framework with the possible combinations
in which seasonality can be accounted for, (2) to establish whether
incorporating seasonality benefits the predictive performance of connectivity
analyses, and (3) to pinpoint at which stages the inclusion of seasonality
provides the largest benefit.

\subsection{African wild dogs in Northern Botswana}

A system well-suited to study the importance of seasonal dynamism for dispersal
and connectivity is the African wild dog population (\textit{Lycaon pictus})
inhabiting the seasonally highly variable Okavango Delta ecosystem in northern
Botswana \citep{McNutt.1996, Wolski.2017}. While once present across the entire
Sub-Saharan continent, the African wild dog has disappeared from a majority of
its historic range due to human persecution, deadly diseases, and habitat
destruction \citep{Woodroffe.2020a}. With about 6,000 adult individuals
remaining in the wild, the species is considered as endangered on the IUCN red
list. Wild dogs are pack-living carnivores, primarily active during the cooler
morning and evening hours \citep{Rasmussen.2012} or during moonlit nights
\citep{Cozzi.2012}. Higher ambient temperature is associated with shorter
activity periods, as the species usually rests during times of elevated heat
\citep{Rabaiotti.2019}. Upon reaching sexual maturity, individuals born into a
pack disperse in single-sex coalitions to find suitable mates and a territory to
settle \citep{McNutt.1996}. In Botswana, the timing of dispersal is seasonal,
with female dispersal peaking prior to the mating season in March, and male
dispersal peaking at the onset of the rainy season in December
\citep{Behr.2020}. Euclidean distances moved by dispersers range from 5 km to
500 km, with some coalitions covering several hundred kilometers within only a
few days \citep{Davies-Mostert.2012, Masenga.2016, Cozzi.2020,
Sandoval-Seres.2022}. Studies of habitat-selection during dispersal show that
wild dogs avoid water, prefer moving along it, prefer moving across open grass
or shrubs, yet avoid areas dominated by humans and densely covered by forests
\citep{ONeill.2020, Hofmann.2021}.

\begin{figure}
 \begin{center}
  \includegraphics[width = \textwidth]{Figures/GraphicalAbstract.pdf}
  \caption{Overview of the different dimensions in which seasonality can be
  rendered in studies of landscape connectivity. (a) During model fitting, the
  modeler needs to decide whether to represent the environment by a static set
  of covariate layers, thus ignoring seasonality, or whether to obtain a dynamic
  set covariate layers that allow rendering it. One also needs to decide
  whether to parameterize a single-season model, assuming fixed preferences
  across the year, or to engage in a multi-season model that accounts for
  seasonal differences. (b) When utilizing the fitted model to predict
  connectivity, one can either assume a static set of environmental covariates
  or again attempt a seasonal take that renders how connectivity differs
  depending on the season. (c) Depending on these decisions, six different
  combinations with differing degrees of dynamism emerge. Our hypothesis was
  that increasing the degree of dynamism would lead to improved predictions.
  However, we were particularly interested in determining at which stage the
  inclusion of seasonally provided the biggest benefits.}
  \label{GraphicalAbstract}
 \end{center}
\end{figure}

\subsection{What We Did}

Here, we examined the importance of including seasonally-varying factors to
assess landscape connectivity for the African wild dog. For this, we compared
the predictive performance of iSSFs that differ in terms of the dynamism they
represented. Specifically, we investigate if and to what degree seasonality at
different stages of the connectivity modeling workflow contributes to improved
predictions. For this, we compile an extensive collection of remote sensed time
series data that accurately render seasonality across the Okavango Delta
ecosystem. We combine them with multi-year dispersal data from
\inputy{GeneralMetrics/CollarsTotal} dispersing wild dog coalitions in northern
Botswana and apply k-fold cross-validation for case-control studies to compare
the predictive efficacy of each model. Finally, we employ IBMMs to simulate
dispersal and estimate connectivity resulting from varying degrees of dynamism.
We hypothesized that habitat-selection and movement behavior would differ
significantly between seasons and that increasing the degree of dynamism would
result in better agreement between observed and predicted dispersal patterns
(\Cref{GraphicalAbstract}).

@@

\subsection{Brief Summary}

>>>>

\documentclass[../FinalThesis.tex]{subfiles}
\begin{document}
\chapter{The Effects of Increasing Seasonal Dynamism when Predicting
Connectivity: Advantages or Unnecessary Complications?}
\label{ChapterSeasonality}
\thispagestyle{empty}
\vspace{-1cm}
\noindent\hfil\rule{0.75\textwidth}{.4pt}\hfil

\begin{center}
David D. Hofmann \orcid{0000-0003-3477-4365},
Dominik M. Behr \orcid{0000-0001-7378-8538},
John W. McNutt,
Arpat Ozgul \orcid{0000-0001-7477-2642}, and
Gabriele Cozzi \orcid{0000-0002-1744-1940}

% \begin{flushleft}

% \vspace{5cm}
\vfill

% \textsuperscript{1} Department of Evolutionary Biology and Environmental
% Studies, University of Zurich, Winterthurerstrasse 190, 8057 Zurich,
% Switzerland.
%
% \vspace{0.25cm}
%
% \textsuperscript{2} Botswana Predator Conservation Program, Wild Entrust,
% Private Bag 13, Maun, Botswana.
%
% \vspace{0.25cm}
%
% \textsuperscript{\S} Corresponding author:
% \href{mailto:david.hofmann2@uzh.ch}{david.hofmann2@uzh.ch}

\end{center}

% \vfill

% \textbf{Running Title:} Connectivity in Seasonal Landscapes
%
% \vspace{0.5cm}
%
% \textbf{Keywords:} African wild dog, Connectivity, Dispersal, Dynamism,
% Individual-based, Lycaon pictus, Okavango Delta, Seasonality

% \end{flushleft}

\newpage
\section*{Abstract}

Seasonally changing conditions can drastically alter landscape connectivity.
Nevertheless, most connectivity studies ignore seasonal dynamism and instead
employ a static set of spatial covariates and assume their focal species to
exhibit a static set of preferences. Ignoring seasonality may, however, mask
important ecological features and processes, thus resulting in poor agreement
between predicted and observed movements and therefore a misrepresentation of
connectivity.

@@

Capitalizing on natural seasonal fluctuations of the Okavango Delta in northern
Botswana and on GPS data collected on dispersing African wild dogs
(\textit{Lycaon pictus}) across different seasons, we investigate the degree to
which a better representation of seasonal dynamism improves our ability to
predict dispersal and connectivity. For this, we fit integrated step-selection
functions and predict connectivity using an individual-based dispersal
simulation while explicitly considering seasonal dynamism in both environmental
covariates and the species' preferences.  Using a rigorous cross-validation
procedure, we compare predictive model performances under each of the six
proposed configurations. While we expected that an increasing degree of seasonal
dynamism would lead to improved predictions, we were particularly interested in
identifying at which stage the inclusion of seasonality would provide the
biggest benefits.

We show that, for our study system, improvements in predictive performances
achieved by incorporating seasonal dynamism were moderate. In fact, accounting
for seasonality only improved predictions when an overly simplistic model
formula assumed. Upon using a more complex model formula, the benefits of
accounting for seasonality vanished, resulting in imperceptible performance
differences. Despite this, patterns of connectivity as obtained from dispersal
simulations revealed marked differences between the most static and most dynamic
configurations. Most notably, connectivity was more homogeneously distributed
throughout the study area when seasonality was taken into account, suggesting
the existence of seasonal stepping stones that facilitate dispersal into
otherwise inaccessible areas.

@@

\newpage
\section{Introduction}

% \subsection{Connectivity}

Landscape connectivity is defined as the degree to which the landscape
facilitates or impedes movement among habitat patches \citep{Taylor.1993} and is
a central prerequisite for maintaining biodiversity \citep{Fahrig.2003}.
Improved connectivity facilitates dispersal \citep{Doerr.2011, Baguette.2013},
which in turn promotes genetic diversity \citep{Perrin.2000, Frankham.2002} and
the colonization of vacant habitats \citep{Hanski.1999, MacArthur.2001}. Due to
its beneficial impacts on metapopulation dynamics, restoring connectivity is
among the most frequently recommended strategies to conserve biodiversity and to
promote resilience against climate change \citep{Heller.2009, Rudnick.2012}.
Quantifying connectivity and identifying critical dispersal corridors have
therefore become crucial tasks in conservation science \citep{Heller.2009,
Rudnick.2012, Keeley.2019, Hofmann.2021}. One aspect that has received limited
attention in the study of connectivity is the role of seasonality. Yet, seasonal
changes in the environment and an animal's willingness to traverse a particular
habitat type can profoundly impact landscape connectivity \citep{Zeller.2020a}.

% \subsection{Seasonality}

Seasonality can impact functional connectivity through spatio-temporal variation
in the landscape itself, or through temporal variation in species' preferences
towards prevailing conditions \citep{Mui.2017, Simpkins.2017a, Zeller.2020a}. In
ecosystems that experience alternations between wet and dry seasons, for
example, the onset of the rainy season initiates distinct ``green-up'' waves,
which affect the availability of food resources for herbivores and,
subsequently, shape their movements \citep{Merkle.2016}. In its most remarkable
form, the variation in environmental conditions drives herbivore migrations
across massive spatial scales, resulting in short-lived movement corridors
between otherwise disconnected habitats \citep{Serneels.2001, Naidoo.2016}.
Alternatively, seasonality can affect functional connectivity via temporal
changes in a species' habitat preferences. Amphibians, for instance, require
both aquatic and terrestrial habitats, but their preference for one habitat over
the other heavily depends on the season \citep{Baldwin.2006}. Although such
seasonal intricacies are likely to play a fundamental role in many ecosystems,
they only rarely enter connectivity studies in an explicit manner. In fact, most
connectivity studies represent their study system by a static set of
environmental layers and assume that their focal species to exhibit a fixed set
of preferences (e.g., \citealp{Elliot.2014, Abrahms.2017, Brennan.2020}).
However, this may result in biased connectivity estimates and a misallocation of
scarce conservation funds \citep{Osipova.2019, Zeller.2020}. Therefore, a more
dynamic approach to connectivity that acknowledges and renders seasonal
variation has been recommended \citep{Zeller.2020a}.

% \subsection{Modeling Functional Connectivity}

Functional connectivity can be estimated using a variety of modeling techniques
\citep{Diniz.2019}, which all comprise four main steps. First, presence data of
the focal species, preferably collected during dispersal (\citealp{Elliot.2014,
Vasudev.2015, Benz.2016}, but see \citealp{Fattebert.2015}), and a set of
spatial covariate layers that are believed or known to be critical determinants
of connectivity are compiled. Second, these data are combined and fed into a
selection model that enables estimating the focal species' preferences towards
environmental features. Popular frameworks for estimating preferences are
case-control designs, where covariate values extracted at observed locations are
contrasted with those extracted at available locations \citep{Beyer.2010,
Fieberg.2010}. This can be achieved using point-selection functions
\citep{Boyce.2002, Manly.2007}, path-selection functions \citep{Cushman.2010},
and step-selection functions \citep{Fortin.2005, Thurfjell.2014}. A particularly
powerful approach is that of \textit{integrated} step-selection functions
(iSSFs), which provides an effective means to jointly model an animal's movement
kernel (i.e., movement capacity or movement preferences) and habitat-selection
\citep{Avgar.2016, Fieberg.2021}. Third, inferred preferences are used to
predict a permeability surface, which is a spatial layer that indicates the
expected ease or difficulty at which the focal species can traverse a certain
area given the area's environmental characteristics \citep{Zeller.2012}.
Finally, in a fourth step, the permeability surface serves as an input to a
connectivity model that reveals crucial movement corridors. At present, the most
popular connectivity models are least-cost path analysis \citep{Adriaensen.2003}
and circuit theory \citep{McRae.2008}, although individual-based movement models
(IBMMs) have recently gained some momentum too \citep{Kanagaraj.2013,
Allen.2016, Hauenstein.2019, Zeller.2020, UnnithanKumar.2022a,
UnnithanKumar.2022, Hofmann.2023}. A major benefit of using IBMMs is that they
allow estimating connectivity directly via simulated dispersal trajectories,
thus bypassing the generation of a permeability surface
\citep{UnnithanKumar.2022a, Hofmann.2023}.

% \subsection{Incorporating Seasonality in Functional Connectivity Models}

Seasonality can enter the connectivity modeling workflow described above at
three distinct stages (\Cref{GraphicalAbstractCH3}). At the first stage, one can
either extract covariate values on environmental features, such as water and
vegetation, from static layers (i.e., a single snapshot per covariate) or from
stacks of layers (i.e., a sequence of snapshots per covariate) that capture
seasonal variation across the landscape (\Cref{GraphicalAbstractCH3}, Stage 1).
The former approach has historically been the norm (e.g., \citealp{Elliot.2014,
Brennan.2020}), yet advances in remote sensing technologies and a facilitated
access to petabytes of landscape data have opened up new avenues for obtaining
spatial layers at unprecedented spatio-temporal resolutions \citep{Toth.2016,
Rumiano.2020}, such that the representation of study systems via dynamic
covariate layers has become more frequent (e.g., \citealp{Osipova.2019,
Kaszta.2021}). At the second stage, one can assume their focal species to
exhibit a fixed set of preferences across seasons by pooling all presence data,
or can try to account for seasonal changes in preferences by splitting the data
accordingly (\Cref{GraphicalAbstractCH3}, Stage 2; e.g., \citealp{Fortin.2005,
Manly.2007, Cushman.2010, Zeller.2020}). \citet{Chetkiewicz.2009}, for instance,
partitioned their data by season to derive seasonal habitat preferences for
pumas (\textit{Puma concolor}) and grizzlies (\textit{Ursus arctos}). At a third
stage, one can estimate connectivity for either ``average'' environmental
conditions, or utilize seasonally updated layers to estimate connectivity for
distinct seasons (\Cref{GraphicalAbstractCH3}, Stage 3). For connectivity models
that rely on permeability surfaces, the latter case implies repeatedly applying
the connectivity model using seasonally updated permeability surfaces (e.g.
\citealp{Osipova.2019, Zeller.2020, Kaszta.2021, Ciudad.2021}). This is
demonstrated \textit{ad extremum} by \citet{Kaszta.2021}, who prepared monthly
updated permeability surfaces and repeatedly employed circuit theory to estimate
dynamic connectivity for African elephants (\textit{Loxodonta africana}). IBMMs,
by contrast, allow for a more elegant solution, as seasonality can be modeled as
the simulated dispersers move. Such an approach has, however, not yet been
followed. Irrespective of the method used, including seasonality into
connectivity analyses is analytically and computationally demanding
\citep{Bishop-Taylor.2018}, raising the question to what degree it should be
considered.

% \subsection{African wild dogs in Northern Botswana}

To study the importance of seasonal dynamism for dispersal and connectivity, we
use an African wild dog population (\textit{Lycaon pictus}) inhabiting the
seasonally highly variable Okavango Delta ecosystem in northern Botswana as a
study system \citep{McNutt.1996, Wolski.2017}. While once present across the
entire Sub-Saharan continent, the African wild dog has disappeared from a
majority of its historic range due to human persecution, deadly diseases, and
habitat destruction \citep{Woodroffe.2020a}. With about 6,000 adult individuals
remaining in the wild, the species is considered as endangered on the IUCN red
list. Wild dogs are pack-living carnivores, primarily active during the cooler
morning and evening hours \citep{Rasmussen.2012} or during moonlit nights
\citep{Cozzi.2012}. Higher ambient temperature is associated with shorter
activity periods, as the species usually rests during times of elevated heat
\citep{Rabaiotti.2019}. Upon reaching sexual maturity, individuals born into a
pack disperse in single-sex coalitions to find suitable mates and a territory to
settle \citep{McNutt.1996}. In Botswana, the timing of dispersal is seasonal,
with female dispersal peaking prior to the mating season in March, and male
dispersal peaking at the onset of the rainy season in December
\citep{Behr.2020}. Euclidean distances moved by dispersers range from 5 km to
500 km, with some coalitions covering several hundred kilometers within only a
few days \citep{Davies-Mostert.2012, Masenga.2016, Cozzi.2020,
Sandoval-Seres.2022}. Studies of habitat-selection during dispersal show that
wild dogs avoid water, prefer moving along it, prefer moving across open
grassland or shrubs, yet avoid areas dominated by humans and densely covered by
forests \citep{ONeill.2020, Hofmann.2021}. Given the species' considerable
mobility and the study area's significant seasonal variability, we deemed the
study-system well-suited to examine the importance of seasonal dynamism on
dispersal and connectivity.

\begin{figure}[htpb]
 \begin{center}
  \includegraphics[width = \textwidth]{Figures/GraphicalAbstract.pdf}
  \caption{Overview of the different stages at which seasonality can be rendered
  in studies of dispersal and landscape connectivity. (a) During model fitting,
  one needs to decide whether to represent the environment by a static set of
  covariate layers, thus ignoring seasonality, or to obtain a dynamic set
  covariate layers that allow showing how the landscape changes over time. One
  also needs to decide whether to parameterize a single-season model, assuming
  fixed preferences across the year, or to engage in a multi-season model that
  accounts for seasonal differences. (b) When utilizing the fitted model to
  predict connectivity, one can either assume a static set of environmental
  covariates or again attempt a seasonal take that renders how connectivity
  differs depending on the season. (c) Depending on these decisions, six
  different combinations that differ in terms of the degree of seasonality they
  encapsulate emerge. Our hypothesis was that increasing the degree of dynamism
  when predicting dispersal and connectivity would lead to improved predictions.
  Notably, we were particularly interested in determining at which stage the
  inclusion of seasonally provided the biggest benefits.}
  \label{GraphicalAbstractCH3}
 \end{center}
\end{figure}

Our goal with this article is to (1) create a framework highlighting the
possible avenues through which seasonality can be incorporated into connectivity
analyses, (2) establish whether incorporating seasonality benefits the
predictive performance of dispersal models, and (3) pinpoint at which stages the
inclusion of seasonality provides the largest benefit. For this, we compare the
predictive performances of six configurations that differ in terms of the
dynamism they represent (\Cref{GraphicalAbstractCH3}b). Specifically, we
investigate if and to what degree seasonality at different stages of the
connectivity modeling workflow contributes to improved predictions. We compile
an extensive collection of remote sensed spatial layers that accurately render
seasonality across the Okavango Delta ecosystem. We combine them with multi-year
dispersal data from \inputy{GeneralMetrics/CollarsTotal} dispersing wild dog
coalitions in northern Botswana and apply k-fold cross-validation for
case-control studies to compare the predictive efficacy under each configuration
(\Cref{GraphicalAbstractCH3}b). Finally, we employ IBMMs to simulate dispersal
and estimate connectivity resulting from varying degrees of dynamism. To
distinguish between different configurations, we use codes that highlight the
level of dynamism represented by each configuration (e.g., SSS =
Static-SingleSeason-Static and DMD = Dynamic-MultiSeason-Dynamic). We
hypothesized that habitat-selection and movement behavior of dispersing wild
dogs would differ significantly between seasons and that increasing the degree
of dynamism would result in better agreement between observed and predicted
dispersal patterns (\Cref{GraphicalAbstractCH3}c).

@@

% \subsection{Brief Summary}
